---
title: "MIDPROJECT CODE"
format: html
editor: visual
---

Importing the CSV File containing Raw Data of Unemployment Rate alone

```{r}
lfs_sexage_raw = read.csv("/Users/anushadusakanti/Documents/GitHub/AnushaDusakanti15.github.io/LFS_SEXAGE_I_R_05032024021852784.csv")
```

Understanding the structure of the data frame

```{r}
str(lfs_sexage_raw)
```

Sub-setting the raw data to select Unemployment Rate data of Japanese Youth

```{r}
unemployment_rate_japan = subset(lfs_sexage_raw, Country == 'Japan' & Age == '15 to 24' & Series == 'Unemployment rate')

unemployment_rate_japan
```

Keeping only the required columns, rounding the Unemployment Rate to 2 digits, renaming columns to appropriate names and resetting the row numbers

```{r}
suppressMessages(library(dplyr))
unemployment_rate_japan = select(unemployment_rate_japan, Sex, Time, Value)
unemployment_rate_japan$Value = round(unemployment_rate_japan$Value, digits = 2)
unemployment_rate_japan <- unemployment_rate_japan %>% 
  rename(Year = Time)

row.names(unemployment_rate_japan) = NULL
```

Selecting only the Unemployment Ice Age Years and resetting the row numbers

```{r}
data1993_2004 = subset(unemployment_rate_japan, Year >= 1993 & Year  <= 2004)
row.names(data1993_2004) = NULL
```

Transposing the Unemployment Ice Age data by pivoting the table from long to wide.

```{r}
suppressMessages(library(tidyr))
transposed_1993_2004 = pivot_wider(data1993_2004, names_from = "Sex", values_from = "Value")
```

Grouping the transposed data by Year and renaming the column containing spaces in between

```{r}
grouped_1993_2004 = transposed_1993_2004 %>%
  group_by(Year)
grouped_1993_2004 = grouped_1993_2004 %>% 
  rename(All.persons = 'All persons')
```

Creating an interactive bar plot with the manipulated data frame to portray the population during Unemployment Ice Age

```{r}
suppressMessages(library(plotly))

bar_plt_1993_2004 = plot_ly(grouped_1993_2004, x = ~Year, y = ~All.persons, type = 'bar', name = 'All persons',
             marker = list(color = 'green')) %>% 
    add_trace(y = ~Men, name = 'Men', 
              marker = list(color = 'blue')) %>%
    add_trace(y = ~Women, name = 'Women', 
              marker = list(color = 'red')) %>%
    layout(xaxis = list(title = "Years", tickmode='linear', tickangle = -90),
           yaxis = list(title = "Youth Unemployment Rate (%)"),
           title = 'Youth Unemployment Rate during Ice Age \n Japan (1993-2004)',
            barmode = 'group')
bar_plt_1993_2004
```

Redesigning the bar plot and creating a interactive line plot to portray the population with separate graphs during Unemployment Ice Age

```{r}
suppressMessages(library(ggplot2))
suppressMessages(library(plotly))
cc <- c("green","blue", "red")

graph <- ggplot(data1993_2004, aes(x = Year, y = Value, fill = Sex, group=1,
                                  text = paste("Sex:", Sex, "<br>Year:", Year, "<br>UR:", Value))) +
  geom_line(color="black") +
  geom_point(shape = 21, size = 2.8) + 
  labs(x = "Year", y = "Youth Unemployment Rate(%)", title = "Youth Unemployment Rate in Japan") +
  facet_wrap(~ Sex) +
  scale_fill_manual(values = cc) +
  labs(fill = "Population") +
  theme_grey() +
  scale_x_continuous(breaks = seq(1993, 2004, by = 1))

ggplotly(graph, tooltip = "text")
```

Importing the CSV file containing Japan's Anual GDP Growth Rate obtained from https://databank.worldbank.org/reports.aspx?source=2&series=NY.GDP.MKTP.KD.ZG&country=JPN#advancedDownloadOptions

```{r}
gdp_annual_1960 = read.csv("/Users/anushadusakanti/Documents/GitHub/AnushaDusakanti15.github.io/Japan_GDP_Annual_Growth.csv")
```

Understanding the structure of the dataset

```{r}
str(gdp_annual_1960)
```

Selecting only the columns required

```{r}
gdp_annual_1960 = select(gdp_annual_1960, Time, Value)
```

Keeping only the meaningful rows

```{r}
gdp_annual = slice(gdp_annual_1960, 9:63)
```

Data Manipulation - converting character columns to numeric, rounding off percentage to 2 digits, and renaming columns appropriately

```{r}
gdp_annual$Time = as.numeric(gdp_annual$Time)
gdp_annual$Value = as.numeric(gdp_annual$Value)
gdp_annual$Value = round(gdp_annual$Value, digits = 2)
gdp_annual = gdp_annual %>% 
  rename(GDP_Annual_Growth = 'Value')
gdp_annual = gdp_annual %>% 
  rename(Year = 'Time')
```

Subsetting the population to 'All persons' to get an overall unemployment outlook from the previously created Unemployment Rate data of Japanese Youth data frame

```{r}
unemployment_all_persons = subset(unemployment_rate_japan, Sex == 'All persons')
```

Data Manipulation - converting character columns to numeric, rounding off percetnage to 2 digits, and renaming columns appropriately

```{r}
unemployment_time_value = select(unemployment_all_persons, Year, Value)
unemployment_time_value = unemployment_time_value %>% 
  rename(Unemployment_Rate = 'Value')
```

Doing an inner join on the GDP data frame and Unemployment Years data frame with the common column 'Year'

```{r}
gdp_unemployment = merge(x = gdp_annual, y = unemployment_time_value, by = 'Year')
gdp_unemployment
```

Creating an interactive line chart to depict the relationship between GDP Growth Rate and Unemployment Rate

```{r}
line_gdp_unemp = plot_ly(gdp_unemployment, x = ~Year) %>% 
    add_trace(y = ~GDP_Annual_Growth, name = 'GDP Annual Growth (in Percentage)', type = 'scatter', mode = 'lines') %>% 
    add_trace(y = ~Unemployment_Rate, name = 'Unemployment Rate (in Percentage)', type = 'scatter', mode = 'lines') %>%
    layout(xaxis = list(title = "Years", tickmode='linear', tickangle = -90),
           yaxis = list(title = "Percentage"),
           title = 'GDP Growth Rate vs Unemployment Rate \n Japan (1968-2022)')
line_gdp_unemp
```

Creating a dataset with unemployment rate of all countries in the year 2022

```{r}
countries_info = subset(lfs_sexage_raw, Age == "15 to 24" & Series == "Unemployment rate" & Time == "2022" & Sex == "All persons")

countries_info <- countries_info[order(countries_info$Value), ]

row.names(countries_info) = NULL
```

Plotting an interactive map with unemployment rate of all countries in the year 2022

```{r}
suppressWarnings({
  color_scale <- colorRamp(c("red", "orange","yellow"))

  min_value <- ifelse(length(countries_info$Value) > 0, min(countries_info$Value, na.rm = TRUE), NA)
  max_value <- ifelse(length(countries_info$Value) > 0, max(countries_info$Value, na.rm = TRUE), NA)

  fig <- plot_ly(countries_info,
                 type = "scattergeo",
                 mode = "markers",
                 locationmode = "country names",
                 locations = ~Country,
                 color = ~Value,
                 text = ~paste("Country: ", Country, "<br>UR: ", Value),
                 colors = color_scale,
                 marker = list(size = 10)) %>%
    layout(title = "Unemployment rate in 2022",
           geo = list(
             showframe = FALSE,
             projection = list(type = 'mercator'),
             bgcolor = "lightblue",  
             showcoastlines = TRUE,  
             showland = TRUE,        
             landcolor = "lightgreen",  
             showocean = TRUE,       
             oceancolor = "lightblue"  
           ))%>%
          colorbar(title = "Unemployment Rate",
             tickvals = seq(min_value, max_value, length.out = 5),
             ticktext = seq(min_value, max_value, length.out = 5))

  fig
})
```
